Depletion capabilities in the OpenMC Monte Carlo particle transport code

### 2020-09-30 manuscript.tex
Address reviewer 2 comments

  Simulations of a pressurized water reactor (PWR) pincell and a sodium-cooled
  fast reactor (SFR) assembly were carried out with OpenMC and Serpent. For both
  problems, the use of a high-fidelity depletion chain results in predictions of
  $k_\text{eff}$ that agree within 20--30 pcm between OpenMC and Serpent.
  Predicted actinide concentrations were found to agree to a fraction of a
  percent, and most fission product concentrations were found to agree within
  1\%. The few cases where larger differences were observed can be attributed
  either to differences in how the energy dependence of fission product yields
  is handled or deficiencies in the nuclear data used. OpenMC simulations of the
  PWR and SFR problems using a simplified 228-nuclide depletion chain
  demonstrate that it achieves accuracy close to that of the full, high-fidelity
  depletion chain.
\end{abstract}

@@

The paper is organized as follows. In \cref{sec:methods}, we describe the
methods and algorithms used to solve the depletion equations, the implementation
within OpenMC, and how nuclear data is used to create depletion chains. Then, in
\cref{sec:results}, we present results for two problems that demonstrate the
accuracy of OpenMC when compared with Serpent~\citep{leppanen2015ane}. In
\cref{sec:conclusions}, we summarize the findings, discuss limitations, and
provide commentary on future avenues of inquiry that may be promising.

@@

The equation that governs the transmutation of nuclides by nuclear reactions and
radioactive decay inside an irradiated environment can be written as
\begin{equation}
  \label{eq:depletion}
  \begin{split}
  \frac{dN_i(t)}{dt} = &\sum\limits_{j=1}^n \left [ \int_0^\infty dE \;
  f_{j \rightarrow i}(E) \sigma_j (E, t) \phi(E,t) + \lambda_{j\rightarrow i}
  \right ] N_j(t) \\ &- \left [\int_0^\infty dE \; \sigma_i (E,t) \phi(E,t) +
  \sum\limits_{j=1}^n \lambda_{i\rightarrow j} \right ] N_i(t), \quad i=1,\dots,n ,
  \end{split}
\end{equation}
where
\begin{equation*}
  \begin{split}
  N_i(t) &\equiv \text{density of nuclide $i$ at time $t$} \\
  \sigma_i(E,t) &\equiv \text{transmutation cross section for nuclide $i$ at energy $E$ and time $t$} \\
  \phi(E,t) &\equiv \text{neutron flux at energy $E$ and time $t$} \\
  f_{j \rightarrow i}(E) &\equiv \text{fraction of transmutation reactions in nuclide $j$ that produce nuclide $i$} \\
  \lambda_{j \rightarrow i} &\equiv \text{decay constant for decay modes in nuclide $j$ that produce nuclide $i$} \\
  n &\equiv \text{total number of nuclides.}
  \end{split}
\end{equation*}
By definition, $f_{i\rightarrow i}$ and $\lambda_{i\rightarrow i}$ are zero so
the summations in \cref{eq:depletion} are effectively over all nuclides except
$i$. Note that we have not included the spatial dependence of the flux or cross
sections. For \cref{eq:depletion} to be accurate, the spatial variation of the
reaction rates over the region that is being depleted should be small. For a
reactor model, this may entail using a unique material in each fuel pin that is
depleted separately.\footnote{OpenMC includes functions for automatically
differentiating materials in cells that are repeated throughout a geometry.}
Ultimately, the user is responsible for deciding what level of spatial
discretization is adequate for the problem at hand.

@@

Rather than choose a single method, OpenMC includes three methods for treating
the energy dependence of FPYs. In the first method, FPY data corresponding to a
specific energy is manually selected. This approach may be appropriate if it is
known that neutrons in the system under consideration have a typical thermal,
fast, or fusion energy spectrum. In the second method, fission rates are tallied
above and below a specified cutoff energy. Then, thermal FPYs are used for all
fissions below the cutoff energy, and fast FPYs are used for all fissions above
the cutoff energy. In the third method, the average energy at which fission
events occur is tallied,
\begin{equation}
  \bar{E} = \frac{\int_0^\infty dE \; E \sigma_f(E)\phi(E)}{\int_0^\infty \sigma_f(E) \phi(E)},
\end{equation}
where $\sigma_f$ is the microscopic fission cross section and $\phi$ is the
neutron flux. An effective FPY is then calculated by linearly interpolating
between tabulated FPYs using the average energy. This approach is effectively
the same as that used by ORIGEN~\citep{gauld2011nt}.

@@

@@

\Cref{fig:sfr-actinides} shows the relative difference in predicted actinide
concentrations between OpenMC and Serpent with probability tables turned on. For
both the full and CASL depletion chains, the relative differences are all within
0.2\%. While this may seem close, the observed difference in $^{238}$U and
$^{239}$Pu concentrations merits special mention. In this model, the fuel
material is at \SI{807}{\kelvin}, but the endf71x library contains cross
sections processed at \SI{600}{\kelvin} and \SI{900}{\kelvin}. Thus, Serpent
ends up using probability tables at \SI{600}{\kelvin} rather than
\SI{807}{\kelvin}, which results in less parasitic absorption in $^{238}$U and
consequently less production of $^{239}$Pu (hence a lower $k_\text{eff}$ than
OpenMC with increasing burnup). Although OpenMC also does not have access to
probability tables at the exact fuel temperature, it statistically interpolates
between the probability tables at \SI{600}{\kelvin} and \SI{900}{\kelvin}.
\begin{figure}[H]
  \centering
  \begin{subfigure}[t]{0.45\textwidth}
    \includegraphics[width=\textwidth]{figures/sfr_actinides_full.pdf}
    \caption{Full depletion chain}
    \label{fig:sfr-actinides-full}
  \end{subfigure}
  \hspace{0.05\textwidth}
  \begin{subfigure}[t]{0.45\textwidth}
    \includegraphics[width=\textwidth]{figures/sfr_actinides_casl.pdf}
    \caption{CASL depletion chain}
    \label{fig:sfr-actinides-casl}
  \end{subfigure}
  \caption{Relative difference in actinide concentrations between OpenMC and
  Serpent for the SFR assembly model at 80, 180, 260, and 360 days with
  probability tables on.}
  \label{fig:sfr-actinides}
\end{figure}

When probability tables are turned off, close agreement is observed between
the predicted actinide concentrations in OpenMC and Serpent, as shown in
\cref{fig:sfr-actinides-nop}. In this case, the actinide concentrations agree to
within 0.01\% at all times during depletion for both the full and CASL depletion
chains.
\begin{figure}[H]
  \centering
  \begin{subfigure}[t]{0.45\textwidth}
    \includegraphics[width=\textwidth]{figures/sfr_actinides_full_nop.pdf}
    \caption{Full depletion chain}
    \label{fig:sfr-actinides-full-nop}
  \end{subfigure}
  \hspace{0.05\textwidth}
  \begin{subfigure}[t]{0.45\textwidth}
    \includegraphics[width=\textwidth]{figures/sfr_actinides_casl_nop.pdf}
    \caption{CASL depletion chain}
    \label{fig:sfr-actinides-casl-nop}
  \end{subfigure}
  \caption{Relative difference in actinide concentrations between OpenMC and
  Serpent for the SFR assembly model at 80, 180, 260, and 360 days with
  probability tables off.}
  \label{fig:sfr-actinides-nop}
\end{figure}

@@

While the focus in this work has primarily been on correctness and accuracy, the
performance of a coupled transport--depletion simulation is also an important
consideration that directly impacts usability. The need to determine reaction
rates for hundreds of nuclides leads to a significant performance degredation
during a transport simulation. Research on methods to reduce the computational
expense of obtaining transmutation reaction rates~\citep{salcedoperez2019mc}
show promise but have yet to be incorporated in a release of OpenMC.

>>>>

  Simulations of a pressurized water reactor (PWR) pincell and a sodium-cooled
  fast reactor (SFR) assembly were carried out with OpenMC and Serpent. For both
  problems, the use of a high-fidelity depletion chain results in predictions of
  $k_\text{eff}$ that agree within 20--30 pcm between OpenMC and Serpent.
  Predicted actinide concentrations were found to agree to a fraction of a
  percent, and most fission product concentrations were found to agree within
  1\%. The few cases where larger differences were observed can be attributed
  either to differences in how the energy dependence of fission product yields
  is handled or deficiencies in the nuclear data used. OpenMC simulations of the
  PWR and SFR problems using a simplified 228-nuclide depletion chain
  demonstrate that it achieves accuracy close to that of the full, high-fidelity
  depletion chain with respect to the studied responses.
\end{abstract}

@@

The paper is organized as follows. In \cref{sec:methods}, we describe the
methods and algorithms used to solve the depletion equations, the implementation
within OpenMC, and how nuclear data is used to create depletion chains. Then, in
\cref{sec:results}, we present results for two problems that demonstrate the
accuracy of OpenMC when compared with Serpent. In \cref{sec:conclusions}, we
summarize the findings, discuss limitations, and provide commentary on future
avenues of inquiry that may be promising.

@@

The equation that governs the transmutation of nuclides by nuclear reactions and
radioactive decay inside an irradiated environment can be written as
\begin{equation}
  \label{eq:depletion}
  \begin{split}
  \frac{dN_i(t)}{dt} = &\sum\limits_{j=1}^n \left [ \int_0^\infty dE \;
  f_{j \rightarrow i}(E) \sigma_j (E, t) \phi(E,t) + \lambda_{j\rightarrow i}
  \right ] N_j(t) \\ &- \left [\int_0^\infty dE \; \sigma_i (E,t) \phi(E,t) +
  \sum\limits_{j=1}^n \lambda_{i\rightarrow j} \right ] N_i(t), \quad i=1,\dots,n ,
  \end{split}
\end{equation}
where
\begin{equation*}
  \begin{split}
  N_i(t) &\equiv \text{density of nuclide $i$ at time $t$} \\
  \sigma_i(E,t) &\equiv \text{transmutation cross section for nuclide $i$ at energy $E$ and time $t$} \\
  \phi(E,t) &\equiv \text{neutron flux at energy $E$ and time $t$} \\
  f_{j \rightarrow i}(E) &\equiv \text{fraction of transmutation reactions in nuclide $j$ that produce nuclide $i$} \\
  \lambda_{j \rightarrow i} &\equiv \text{decay constant for decay modes in nuclide $j$ that produce nuclide $i$} \\
  n &\equiv \text{total number of nuclides.}
  \end{split}
\end{equation*}
By definition, $f_{i\rightarrow i}$ and $\lambda_{i\rightarrow i}$ are zero so
the summations in \cref{eq:depletion} are effectively over all nuclides except
$i$. Also note that the transmutation cross section $\sigma_i$ represents the
sum of the cross sections for all reaction channels that result in
transmutation. We have not included the spatial dependence of the flux or cross
sections. For \cref{eq:depletion} to be accurate, the spatial variation of the
reaction rates over the region that is being depleted should be small. For a
reactor model, this may entail using a unique material in each fuel pin that is
depleted separately.\footnote{OpenMC includes functions for automatically
differentiating materials in cells that are repeated throughout a geometry.}
Ultimately, the user is responsible for deciding what level of spatial
discretization is adequate for the problem at hand.

@@

Rather than choose a single method, OpenMC includes three methods for treating
the energy dependence of FPYs. In the first method, FPY data corresponding to a
specific energy is manually selected. This approach may be appropriate if it is
known that neutrons in the system under consideration have a typical thermal,
fast, or fusion energy spectrum. In the second method, fission rates are tallied
above and below a specified cutoff energy. Then, thermal FPYs are used for all
fissions below the cutoff energy, and fast FPYs are used for all fissions above
the cutoff energy. In the third method, the average energy at which fission
events occur is tallied,
\begin{equation}
  \bar{E} = \frac{\int_0^\infty dE \; E \sigma_f(E)\phi(E)}{\int_0^\infty dE \; \sigma_f(E) \phi(E)},
\end{equation}
where $\sigma_f$ is the microscopic fission cross section and $\phi$ is the
neutron flux. An effective FPY is then calculated by linearly interpolating
between tabulated FPYs using the average energy. This approach is effectively
the same as that used by ORIGEN~\citep{gauld2011nt}.

@@

Although both OpenMC and Serpent use CRAM for matrix exponential solutions, they
employ different forms of CRAM. OpenMC relies on the IPF form of CRAM whereas
Serpent relies on the partial fraction decomposition (PFD)
form~\citep{pusa2016nse}. Although the IPF form has been shown to be superior
for long timesteps in decay calculations, both the PFD and IPF form are very
accurate for burnup calculations, and we do not expect differences between the
OpenMC and Serpent results to be attributable to the form of CRAM used. At the
time of writing, OpenMC uses IPF CRAM of order 48 by default and Serpent uses
PFD CRAM of order 14.

@@

\Cref{fig:sfr-actinides} shows the relative difference in predicted actinide
concentrations between OpenMC and Serpent with probability tables turned on. For
both the full and CASL depletion chains, the relative differences are all within
1\%. While this may seem close, the observed difference in $^{238}$U and
$^{239}$Pu concentrations merits special mention. In this model, the fuel
material is at \SI{807}{\kelvin}, but the endf71x library contains cross
sections processed at \SI{600}{\kelvin} and \SI{900}{\kelvin}. Thus, Serpent
ends up using probability tables at \SI{600}{\kelvin} rather than
\SI{807}{\kelvin}, which results in less parasitic absorption in $^{238}$U and
consequently less production of $^{239}$Pu (hence a lower $k_\text{eff}$ than
OpenMC with increasing burnup). Although OpenMC also does not have access to
probability tables at the exact fuel temperature, it statistically interpolates
between the probability tables at \SI{600}{\kelvin} and \SI{900}{\kelvin}. Also
note that $^{239}$U and $^{239}$Np, which are also produced as a result of
neutron capture in $^{238}$U, have a higher relative difference because of their
short half-lives and correspondingly small concentrations.
\begin{figure}[H]
  \centering
  \begin{subfigure}[t]{0.45\textwidth}
    \includegraphics[width=\textwidth]{figures/sfr_actinides_full.pdf}
    \caption{Full depletion chain}
    \label{fig:sfr-actinides-full}
  \end{subfigure}
  \hspace{0.05\textwidth}
  \begin{subfigure}[t]{0.45\textwidth}
    \includegraphics[width=\textwidth]{figures/sfr_actinides_casl.pdf}
    \caption{CASL depletion chain}
    \label{fig:sfr-actinides-casl}
  \end{subfigure}
  \caption{Relative difference in actinide concentrations between OpenMC and
  Serpent for the SFR assembly model at 80, 180, 260, and 360 days with
  probability tables on.}
  \label{fig:sfr-actinides}
\end{figure}

When probability tables are turned off, close agreement is observed between
the predicted actinide concentrations in OpenMC and Serpent, as shown in
\cref{fig:sfr-actinides-nop}. In this case, the actinide concentrations agree to
within 0.04\% at all times during depletion for both the full and CASL depletion
chains.
\begin{figure}[H]
  \centering
  \begin{subfigure}[t]{0.45\textwidth}
    \includegraphics[width=\textwidth]{figures/sfr_actinides_full_nop.pdf}
    \caption{Full depletion chain}
    \label{fig:sfr-actinides-full-nop}
  \end{subfigure}
  \hspace{0.05\textwidth}
  \begin{subfigure}[t]{0.45\textwidth}
    \includegraphics[width=\textwidth]{figures/sfr_actinides_casl_nop.pdf}
    \caption{CASL depletion chain}
    \label{fig:sfr-actinides-casl-nop}
  \end{subfigure}
  \caption{Relative difference in actinide concentrations between OpenMC and
  Serpent for the SFR assembly model at 80, 180, 260, and 360 days with
  probability tables off.}
  \label{fig:sfr-actinides-nop}
\end{figure}

@@

While the focus in this work has primarily been on correctness, the performance
of a coupled transport--depletion simulation is also an important consideration
that directly impacts usability. The need to determine reaction rates for
hundreds of nuclides leads to a significant performance degredation during a
transport simulation. Research on methods to reduce the computational expense of
obtaining transmutation reaction rates~\citep{salcedoperez2019mc} show promise
but have yet to be incorporated in a release of OpenMC.
